---
title: "postAD: RTMB-based fitting for discrete-trait phylogenetic models"
author: "Haoyu Ji"
format:
  html:
    toc: true
    toc-depth: 3
    df-print: paged
    self-contained: true
    number-sections: true
    code-tools: true
    code-fold: show
    code-summary: "Code"
  pdf:
    number-sections: true
    fig-cap: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
vignette: |
  %\VignetteIndexEntry{postAD: RTMB-based fitting for discrete-trait phylogenetic models}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| code-fold: true
#| code-summary: "packages"
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)

library(Matrix)
library(ape)
library(tidyverse); theme_set(theme_bw())
library(expm)
library(RTMB)
library(colorspace)

source(here::here("R", "postAD.R"))
source(here::here("R", "general.R"))
```

# Background information

`postAD` fits continuous-time Markov models for discrete trait evolution on a phylogeny using an RTMB backend. The package is designed around two complementary interfaces:

- **rates interface**: where transition-rate parameters are specified directly
- **formula interface**: where structured hypotheses about transition rates are expressed through formulas.

Both interfaces target the same underlying pruning likelihood. They differ only in how the transition model is parameterized.

## From multiple traits to a single state space

Suppose we have a single binary trait with states 0 and 1. The corresponding transition-rate matrix is
$$
Q=\left[\begin{matrix}
- & q_{1 \to 0}\\
q_{0 \to 1} & -
\end{matrix}\right]
$$
.

This matrix is read from row to column: the entry in row $i$, column $j$ gives the instantaneous transition rate from state $i$ to state $j$.

If we instead model two binary traits jointly, the observed state space becomes
$$
{00,01,10,11},
$$

so that the model is defined on a larger state space. In most applications, we disallow simultaneous changes in more than one trait, so only single-step transitions are permitted.

The main idea behind `postAD` is the same: multiple observed traits are converted into a joint state space, and a continuous-time Markov model is fitted on that state space.

## Two interfaces, one likelihood

The **rates interface** is most useful when the user wants direct control over the free parameters in the transition model.

The **formula interface** is most useful when the user wants to express structured hypotheses compactly. For example, a user may want the transition process for one trait to depend on the states of other traits through additive or interaction terms.

In both cases, the fitted object is obtained by optimizing the same pruning-based objective.

## Data conventions

Throughout this vignette, we assume that:

- `tree` is an object of class `phylo`,
- `traitMatrix` is a data frame,
- The first column of `traitMatrix` contains species names, and all remaining columns are numeric trait columns.
- The species names in `traitMatrix` must match `tree$tip.label` exactly and in the same order, unless the relevant preprocessing step reorders them explicitly.

State vectors are typically specified as `rep(2, n)`, but non-binary traits and combinations of traits with different numbers of states are also allowed, for example `rep(3, n)` or `c(2, 3, 4)`. When a single state value is supplied, it is interpreted as applying to all traits. If no state vector is provided, all traits are assumed to be binary by default.

## Dateset description

The datasets `c_tree` and `c_traits` form a matched phylogeny–trait dataset used throughout this vignette.

`c_tree` is a phylo object containing 234 terminal taxa and 233 internal nodes. Branch lengths are included, and the tree has been reordered in pruningwise order to facilitate pruning-based likelihood calculations.

`c_traits` is a 234 × 4 data frame containing one row per species. The column Species gives the species names, and the remaining columns represent three binary traits:

- ag: accessory gland presence/absence
- care: parental care
- spawning: spawning mode

The ordering of species in `c_traits` has been synchronized with `c_tree$tip.label`, so the two objects can be passed directly into the fitting functions without additional matching or reordering.

# Default use of postAD

## A minimal example

We begin with a small example using a tree and a trait matrix with three binary traits.

```{r datareveal}
c_tree <- readRDS(here::here("data", "c_tree.rds"))
c_traits <- readRDS(here::here("data", "c_traits.rds"))
head(c_traits)
```

A basic postAD fit only requires a tree, a trait matrix, and the number of states per trait:

```{r fit0}
fit0 <- postAD(tree = c_tree, traitMatrix = c_traits, state = 2)
```

Printing the object should display a compact summary of the fitted model:

```{r fit0res}
names(fit0)
```

At minimum, the fit is expected to contain:

- the best parameter vector,
- the optimized objective value,
- the gradient at the optimum
- timing information.

## Understanding the return value

A typical fitted object contains the following key components:

- `pars.best`: the best-fitting parameter vector on the optimization scale;
- `obj.best`: the optimized objective value;
- `gr.best`: the gradient evaluated at pars.best;
- `time_total`: total elapsed time for the fitting routine.

For example:

```{r fit0res2}
str(fit0, max.level = 1)
```

If the model was fitted with `keep_all = TRUE`, the returned object may also include a `result_frame` storing the results of all optimization attempts.

```{r fit0ms}
fit0_ms <- postAD(tree = c_tree, traitMatrix = c_traits, state = 2, keep_all = TRUE)
names(fit0_ms)
head(fit0_ms$result_frame)
```

The optimizer's run-time is also stored in `result_frame`; to view it, call `result_frame$time_opt`.

## Interpreting fit quality

A useful first diagnostic is the gradient norm,
$$
\|\nabla f(\hat{\theta})\|^2 = \sqrt{\sum_{i} \left( \frac{\partial f}{\partial \theta_i} \right)^2} 
$$
In code:

```{r fit0gr}
sqrt(sum(fit0$gr.best^2))
```

Smaller values indicate that the optimizer has reached a point closer to a stationary point of the objective. In practice, the gradient norm should always be interpreted together with:

- the optimizer convergence code,
- the objective value,
- the stability of the solution across restarts, and
- the biological plausibility of the fitted rates.

# The formula interface

## Why a formula interface?

Direct rate parameterization can become cumbersome when the state space is large or when the user wants to express structured hypotheses. For example, it may be natural to say:

- the transition process for `care` is constant;
- the transition process for `spawning` is constant
- the transition process for `ag` depends on `care`, `spawning`, and their interaction.

The formula interface provides a compact way to express such models.

##  A complete formula example

A simple formula-based model can be written as:

```{r formula}
fit_formula <- postAD(tree = c_tree, traitMatrix = c_traits, state = 2, 
                      formula_list = list(ag ~ care * spawning, care ~ 1, spawning ~ 1))
```

This says:

- the transition process for `ag` depends on `care`, `spawning`, and their interaction
- the transition process for `care` has only an intercept term
- the transition process for `spawning` has only an intercept term

## How formulas map to the underlying model

Internally, the formula interface is translated into a parameterized transition model. The exact mapping depends on the package internals, but conceptually the workflow is:

- construct the joint state space
- determine which transitions are allowed
- build design matrices implied by the formulas
- map the resulting coefficients to the transition-rate model

If the package exposes a preprocessing helper such as `Q_prep()`, it can be helpful to inspect it directly:

```{r qprep}
qp <- Q_prep(mode = "formula", nstate = 2, ntrait = 3,
            formula_list = list(ag ~ care * spawning, care ~ 1, spawning ~ 1))
qp$par_names
```

This is often the easiest way to understand exactly which coefficients are being estimated.

## Comparing the rates and formula interfaces

Because both interfaces target the same likelihood, it is natural to compare them on the same dataset.

```{r twofits}
fit_rates <- postAD(tree = c_tree, traitMatrix = c_traits)
fit_formula <- postAD(tree = c_tree,traitMatrix = c_traits,
                      formula_list = list(ag~care*spawning,care ~ 1,spawning ~ 1))
data.frame(
  method = c("rates", "formula"),
  objective = c(fit_rates$obj.best, fit_formula$obj.best),
  grad_norm = c(sqrt(sum(fit_rates$gr.best^2)),sqrt(sum(fit_formula$gr.best^2))),
  total_time = c(fit_rates$time_total, fit_formula$time_total)
)
```

When comparing the two interfaces, it is important to distinguish between:

- equality of the optimized objective value,
- similarity of the fitted parameter vectors, and
- similarity of the implied transition model.

Depending on the parameterization, raw parameter vectors may not always be directly comparable term by term even when the implied fitted model is similar.

# Optimization, diagnostics, and reproducibility

## Starting values

If `pars = NULL`, `postAD` generates an initial parameter vector automatically

```{r aotopars}
fit_auto <- postAD(tree = c_tree, traitMatrix = c_traits)
```

Users may also supply their own starting values explicitly:

```{r manualpars}
pars0 <- rep(log(0.1), 24)
fit_user <- postAD(tree = c_tree,traitMatrix = c_traits,pars = pars0)
```

Custom starting values are particularly useful for difficult models or when comparing alternative optimization strategies.

## Multistart fitting

For complex objectives, a single optimization run may converge to a poor local optimum. `postAD` therefore supports multistart fitting:

```{r onecore}
onecore <- system.time(
  fit_ms <- postAD(tree = c_tree, traitMatrix = c_traits,
    multistart = 10,jitter.sd = 0.5,keep_all = TRUE)
)
```

The `result_frame` returned by `keep_all` contains results of each starts:

```{r keepall}
head(fit_ms$result_frame)
```

This allows the user to inspect:

- objective values across starts
- convergence codes
- timing information
- the stability of the final solution

## Parallel execution

Multistart fitting can be parallelized:

```{r ninecore}
ninecore <- system.time(
  fit_ms_par <- postAD(tree = c_tree, traitMatrix = c_traits,
    multistart = 10, parallel = TRUE, keep_all = TRUE)
)
onecore; ninecore
```

Parallel execution is most useful when:

- the objective is expensive to evaluate
- many random restarts are desired
- multiple CPU cores are available

Just as the above example shown, if the task is not sufficiently complex, serial computation is generally still recommended.

## Reproducibility

`postAD` uses random numbers in at least two contexts:

- generating initial values when `pars = NULL`
- generating perturbed starts when `multistart > 1`

For reproducible analyses, users should set a seed explicitly:

```{r rep}
fit_rep <- postAD(tree = c_tree,traitMatrix = c_traits, seed = 101)
fit_rep$pars.best
```

When benchmarking or simulation and fitting are combined in the same workflow, it is often preferable to use separate seeds for data generation and model fitting.

# Advanced workflows

## Returning low-level objects

Advanced users may want direct access to the underlying RTMB objective object or a lower-level function interface.

```{r obj}
obj <- postAD(tree = c_tree,traitMatrix = c_traits,return_obj = TRUE)
str(obj)
```

If a raw interface is supported:

```{r showobj}
raw_obj <- postAD(tree = c_tree, traitMatrix = c_traits,
  return_obj = TRUE,return_obj_type = "raw")
str(obj)
```

This can be useful for:

- custom optimizers
- profiling
- debugging
- benchmarking
- Hessian diagnostics
- downstream methodological development

## Custom optimization workflows

Once a low-level objective is available, users can fit the model using their own optimization wrapper.

```{r optobj}
obj <- postAD(tree = c_tree, traitMatrix = c_traits, return_obj = TRUE)
fit_custom <- nlminb(start = obj$par, objective = obj$fn,gradient = obj$gr)
fit_custom
```

This is particularly useful when comparing optimizers or when embedding postAD into larger workflows.

## Common failure modes

The most common causes of failure are data-format problems rather than numerical issues. For example:

- `traitMatrix` contains more than one non-numeric column
- species names in `traitMatrix` do not match `tree$tip.label`
- trait columns are not numeric after import
- the specified `state` does not match the coding in the data
- the model is overparameterized relative to the data

## What to compare across fits

When comparing alternative model specifications, the most useful summaries are typically:

- optimized objective value
- gradient norm
- elapsed time
- convergence code
- stability across starts
- the implied fitted transition model

Direct comparison of raw parameter vectors is most meaningful when the two fitted models share the same parameterization.

# Summary

`postAD` provides an RTMB-based framework for fitting discrete-trait phylogenetic models on a joint state space. The package supports both a direct rates interface and a higher-level formula interface. These interfaces target the same underlying likelihood while providing different levels of modeling convenience and control.

For most users, the recommended workflow is:

- begin with the default interface on a small example
- inspect the fitted object and the gradient norm
- move to the formula interface when structured hypotheses are needed
- use multistart fitting and low-level object return only when necessary

# References

Boyko, J. D., and Beaulieu, J. M. corHMM 2.1: Generalized hidden Markov models.

RTMB documentation and package references.